Depression and suicidality as costly signals of need

### 2021-04-27 Paper.rmd
Continuing to work on intro

@@

Withholding or risking one's contributions to a cooperative endeavor can be conceptualized as credibly signaling private information about one's need. Cooperation generates benefits over time. Thus, withholding or risking contributions incurs a *cost of delay*. In a classic non-cooperative game theory model of bargaining over the division of benefits, @Rubenstein:1982 showed that two parties can come to an immediate agreement over division of benefits despite conflicts of interest if the parties' valuations of cooperation are not private information. If valuations are private information, however, then immediate agreement might not be possible because each party has an incentive to exaggerate their need. A willing to delay, i.e., incur a cost, is a credible signal that one has a low valuation of the endeavor, and is therefore genuinely in need. Once valuations are known, the parties can reach an agreement on the division of benefits [@KennanWilson1993; @feinberg2005].

@@

<!-- Likely the most reliable finding regarding depression and suicidality is that most cases are closely tied to major negative life events [@Hammen2005; @Mazure1998; @devries2011; @ellsberg2008]. Compared to non-depressed individuals, those with depression tend to report at least twice as many instances of negative events than non-depressed individuals [@Mazure1998] and more negative events than those with schizophrenia and bipolar depression across multiple studies [@Paykel1994]. In longitudinal studies, the onset of depression has been reported to quickly follow the negative event [@KendlerEtAl1999] or coincide with periods where conflict is likely to increase prior to it (e.g., depression starting before a divorce rather than after; @rosenstrom2017).  -->

--\>

<!-- This strong connection between negative life events and depression along with the adversity and distress that likely accompanies these events has led to the suggestion that this relationship is causal [@KendlerEtAl1999]. Support for this comes from the fact that the relationship between negative events and depression holds true when looking solely at events outside of one's control, showing that the connection is unlikely to be driven solely by individuals who are already depressed creating situations where negative events are likely to be common [@Hammen2005; @KendlerEtAl1999]. In addition to this, twin studies have shown that one's history of negative events is still a strong predictor of major depression when controlling for genetic similarity and that part of the heritability of depression stems from the heritability of negative events, like divorce and family conflict [@KendlerBaker2007; @KendlerEtAl1999]. -->

<!-- Likely the most reliable finding regarding depression and suicidality is that most cases are closely tied to major negative life events [@Hammen2005; @Mazure1998; @devries2011; @ellsberg2008], with such events likely causing depressive states based on twin studies and examinations involving negative events outside of one's control[@Hammen2005; @KendlerEtAl1999]. Of the types of negative life events associated with depression, those which involve conflict tend to be more strongly associated with depression [@Hammen2005; @Mazure1998]. Marital problems, bullying, and the existence of abusive relationships are all commonly cited risk factors for depression [@KendlerEtAl1995; @KendlerEtAl1999; @klomek2007; @klomek2019], with assault (both sexual and non-sexual), in particular, greatly increasing one's risk of depression [@KendlerEtAl1995; @KendlerEtAl1999]. Other notable risk factors like loss of a loved one, financial problems, job loss, or severe or prolonged illness, might seem less related to conflict. However, the ability for others to provide meaningful support to limit the harm caused in times of need [@SugiyamaSugiyama2003] combined with the ubiquity of conflicts of interests mean that many types of problems will eventually involve social conflict even if they did not start that way [@Hagen2003]. In line with this reasoning, conflict, as a whole, was found to be associated with depression among the Tsimane, particularly when it involved non-kin [@StieglitzEtAl2015a]. -->

<!-- Due to depression's relationship to sucide [@who2001; @hawton2013], it should not be surprising that suicidality also often coincides with social adversity and conflict. Grief over the loss of a loved one [@firth1936: @Firth:1961; @devries2011], bullying [@alavi2017; @butler2019; @ford2017; @wang2020], and instances of social transgression [@Firth:1961; @Malinowski:1932; @SymeHagen2018; @lester1997] have all been reported to lead to suicidal behavior in both industrialized and traditional societies. Of the recorded instances of suicide attempts in the Human Relations Area Files (HRAF), 77.4% of attempts were found to be linked to times of social conflict, 88.7% were linked to fitness threats in general, and 84.9% were linked to situations in which attempters were powerless to resolve their situation alone [@SymeEtAl2016]. As with depression, assault is an particularly powerful risk factor for suicidal behavior based on multiple large-scale multi-country studies [@devries2011; @ellsberg2008] and more localized examinations [@reigstad2017]. -->

>>>>

There is strong evidence that most cases of depression are caused by major negative life events [@Hammen2005; @Mazure1998; @devries2011; @ellsberg2008]. Compared to non-depressed individuals, those with depression report about twice as many instances of negative events than non-depressed individuals [@Mazure1998] and more negative events than those with schizophrenia and bipolar depression across multiple studies [@Paykel1994]. In longitudinal studies, depression onsets soon after a negative event [@KendlerEtAl1999] or coincides with periods where adversity is likely to increase prior to it [e.g., depression starting before a divorce rather than after; @rosenstrom2017]. Additional support for the causal role of negative life events comes from the fact that the relationship between negative events and depression holds true when looking at events outside of one's control, showing that the connection is unlikely to be driven solely by individuals who are already depressed creating situations where negative events are likely to be common [@Hammen2005; @KendlerEtAl1999]. Additionally, twin studies have shown that one's history of negative events is still a strong predictor of major depression when controlling for genetic similarity, and that part of the heritability of depression stems from the heritability of negative events, like divorce and family conflict [@KendlerBaker2007; @KendlerEtAl1999].

There is equally strong evidence that depression is closely associated with anger and conflict [@cassiello2016]. Of the types of negative life events associated with depression, those which involve conflict tend to be more strongly associated with depression [@Hammen2005; @Mazure1998]. Marital problems, bullying, and abusive relationships are all common risk factors for depression [@KendlerEtAl1995; @KendlerEtAl1999; @klomek2007; @klomek2019], with assault (both sexual and non-sexual), in particular, greatly increasing one's risk of depression [@KendlerEtAl1995; @KendlerEtAl1999]. This holds true even in a small-scale, non-Western society: among the Tsimane, Amazonian horticulturalists, depression is also associated with conflict, particularly conflict involving non-kin [@StieglitzEtAl2015a]. See @SymeHagen2021 for a review of the association of depression with anger and conflict.

Other notable risk factors like loss of a loved one or severe or prolonged illness, might seem less related to conflict. In these situations, however, individuals will often need increased investment from social partners, who might not be able, or want, to provide more investment than they already are. The ability for others to provide meaningful support to limit the harm in times of need [@SugiyamaSugiyama2003] combined with the ubiquity of conflicts of interests mean that many types of problems will eventually involve social conflict even if they did not start that way [@Hagen2003]. There is indeed increasing evidence that loss of a loved one is often followed by increased family conflict [see @SymeHagen2021 for a brief review].

# Bargaining: Credibly signaling need under conflict

Individuals who have suffered a negative life event will often need help to mitigate the fitness cost. To obtain this help, they must signal need, e.g., by verbally requests, sad facial expressions, and crying. But if they have conflicts of interest with social partners who could provide that help, and if their need is private information, their signals might not be believed if the signals, like crying, could easily be faked [@Zahavi1975; @mokkonen2016] and if the cost of believing a deceptive signal was high [@dawkins1978; @johnstone1993; @Hasson1994].

demonstrating that one's need is genuine has likely been a recurrent adaptive problem since humans evolved the capacity to provide instrumental support to one another.

@@

Withholding or risking one's contributions to a cooperative endeavor can be conceptualized as credibly signaling private information about one's need. Cooperation generates benefits over time. Thus, withholding or risking contributions incurs a *cost of delay* -- the more one values the endeavor, the greater the cost of delay. In a classic non-cooperative game theory model of bargaining over the division of benefits, @Rubenstein:1982 showed that two parties can come to an immediate agreement over division of benefits despite conflicts of interest if the parties' valuations of cooperation are not private information: in this case, each party knows exactly what division of benefits the other will accept, and can therefore make that offer immediately, avoiding the cost of delay. If valuations are private information, however, then immediate agreement might not be possible because each party has an incentive to deceptively report a low valuation in the hopes of getting more. A willingness to delay is a credible signal that one has a low valuation of the endeavor, and is therefore genuinely in need, because for those with a low valuation the benefit of a better offer outweighs the low cost of delay whereas for those for those with a high valuation the benefit of a better offer does not outweigh the higher cost of delay. Once valuations are known, the parties can reach an agreement on the division of benefits [@KennanWilson1993; @feinberg2005].

@Rosenthal1993 proposed a credible signaling model of what he termed "gambling-type" suicide attempts in which a probability of death less than 1 serves as the signal cost that allows signal receivers to accurately discriminate between those who are genuinely in need and those who are not.

@@

<!-- Due to depression's relationship to suicide [@who2001; @hawton2013], it should not be surprising that suicidality also often coincides with social adversity and conflict. Grief over the loss of a loved one [@firth1936: @Firth:1961; @devries2011], bullying [@alavi2017; @butler2019; @ford2017; @wang2020], and instances of social transgression [@Firth:1961; @Malinowski:1932; @SymeHagen2018; @lester1997] have all been reported to lead to suicidal behavior in both industrialized and traditional societies. Of the recorded instances of suicide attempts in the Human Relations Area Files (HRAF), 77.4% of attempts were found to be linked to times of social conflict, 88.7% were linked to fitness threats in general, and 84.9% were linked to situations in which attempters were powerless to resolve their situation alone [@SymeEtAl2016]. As with depression, assault is an particularly powerful risk factor for suicidal behavior based on multiple large-scale multi-country studies [@devries2011; @ellsberg2008] and more localized examinations [@reigstad2017]. -->
